# Simple Linear Regression {#sec:slr}

<!--- For HTML Only --->
`r if (!knitr:::is_latex_output()) '
$\\newcommand{\\E}{\\mathrm{E}}$
$\\newcommand{\\Var}{\\mathrm{Var}}$
'`

```{r include=FALSE}
library(tidyverse)
library(palmerpenguins)
library(broom)
```

<!-- ## Overview -->

## The Simple Linear Regression (SLR) Model

### Goal 

In simple linear regression (SLR), our goal is to find the best-fitting straight line, commonly called the **regression line**, through a set of paired $(x, y)$ data. 

For example, we could seek the best-fitting straight line for the penguin data:

```{r g-penguin-flip-mass-lm1b, echo=F, message=FALSE}
g_penguin_flip_mass <- penguins %>%
  # rename(Species=species) %>%
  filter(!is.na(body_mass_g)) %>%
  ggplot(aes(x=flipper_length_mm,
           y=body_mass_g)) +
  theme_bw() + 
  geom_point(aes()) +
    xlab("Flipper Length (mm)") +
    ylab("Body Mass (g)")
g_penguin_flip_mass_lm <- g_penguin_flip_mass +
  geom_smooth(se=FALSE, method="lm", col="black")
g_penguin_flip_mass_lm
```
 
Throughout this chapter and the next, we will look more at how to interpret what this line means and how to estimate it.

### SLR model

The equation for the **Simple Linear Regression (SLR) model** is:

\begin{equation}
Y_i = \beta_0 + \beta_1x_i + \epsilon_i
(\#eq:slr)
\end{equation}

where:

* $Y_i$ is a *random variable* representing the outcome (e.g. body mass)

* $x_i$ is a fixed predictor variable (e.g., flipper length)

* $i$ is an index for each observation (penguins, people, units, etc.)

* $\beta_0$ is a parameter representing the intercept

* $\beta_1$ is a parameter representing the slope

* $\epsilon_i$ is a *random variable* representing variation (the "error" in the model)

These five elements can be categorized as three different kinds of quantities:

* **Random variables** ($Y_i$ and $\epsilon_i$): Random quantities that are not directly observed. For a given dataset, the corresponding observed values will be denoted $y_i$ and $e_i$ (See Section \@ref(sec:realtheo))
* **Parameters** ($\beta_0$ and $\beta_1$): Numbers that are fixed but unknown. These values are the same for all observations in the model. Estimating the values of these parameters (the topic of Section \@ref(slrestimation)) is the basic task in "fitting" a regression model. 
* **Observed data** ($x_i$): Values that are fixed and known. Each observation can have a different value of $x_i$, or there could be as few as only two distinct values for $x_i$. Even though these might vary between datasets, the SLR model considers them fixed.
SLR refers specifically to settings where there is only one predictor variable. In Chapter \@ref(mlr) we will extend this to multiple predictor variables.


### SLR Model Assumptions

The assumptions corresponding to the SLR model (equation \@ref(eq:slr)) are:

1. $E[\epsilon_i] = 0$
   * The error terms have mean zero. This allows the $\epsilon_i$ to account for variability above and below the line.
2. $Var[\epsilon_i] = \sigma^2$ 
   * The error terms have constant variance. (If the variance was different for each observation, then $\Var[\epsilon_i] = \sigma_i^2$.)
3. $\epsilon_i$ are uncorrelated
   * Each observation gives you new information. This usually means that each observation is independent.

<!-- **Note:** there is no assumption about a normal distribution! -->

For now, the key assumption is $E[\epsilon_i] = 0$. Importantly, we do not require an assumption about normality of the error term. 
 We will re-visit  these  assumptions in detail, and discuss what happens when they are violated, in Chapter \@ref(sec:modadequacy). 


## Parameter Interpretation

### Regression Line Equation

A consequence of Sssumption 1 ($\E[\epsilon_i] = 0$) is that the mean of $Y_i$ is a straight line:

\begin{align}
\E[Y_i] &= \E[\beta_0 + \beta_1x_i + \epsilon_i] \notag\\
&=  \E[\beta_0 + \beta_1x_i] + \E[\epsilon_i]\notag\\
&=  \beta_0 + \beta_1x_i + 0 \notag\\
&=  \beta_0 + \beta_1x_i \label{eq:slrmean}
\end{align}

This allows us to interpret the parameters $\beta_0$ and $\beta_1$ in terms of the  slope and intercept of the line.


### Interpreting $\beta_1$
 
The $\beta_1$ parameter represents the **slope** of the regression line. In other words, $\beta_1$ is the difference in $\E[Y_i]$ between observations that differ in $x_i$ by one unit.

In almost any data analysis project, an important task is providing an interpetation of the model parameters. When describing the interpretation of $\beta_1$, it is important to include these key elements:

* An interpretation of $\beta_1$ as an **estiumated difference** in the average value of the outcome variable
  * For linear regression, "difference in average" is the same as "average difference". You can use either phrase.
  * In this section only, we are assuming that $\beta_1$ is known. Starting in Section \@ref(slrestimation) we will introduce estimates of this parameter and start including "estimated" in interpretations.
* Specify that this is for a 1-unit difference in the predictor variable (x).
* Be cautious about referring to $\beta_1$ as an "increase" or a "decrease". Those words imply that an intervention was conducted, in which the value of $x$ was directly modified. This is possible in controlled experiments, but is not the case for observational datasets.
* Include confidence intervals (see Section \@ref(slrCI))^[The examples in this chapter will skip this for now. This will be added in the examples in Section \@ref(slrCI)]
* If appropriate, include information about the conclusion from a hypothesis test (see Section \@ref(slrhypothtest))^[The examples in this chapter will skip this for now. This will be added in the examples in Section \@ref(slrhypothtest)]
* Include units for both the outcome and predictor
* If known, include statement about population/context.


```{example peng-beta1}
Consider two groups of penguins:
```

* Penguins that have flipper lengths of 201mm. Call this "Group A"
* Penguins that have flipper lengths of 200mm. Call this "Group B"

Assuming the SLR model \@ref(eq:slr), what is the difference in average body mass between these groups of penguins? To answer this, first write out the regression equation for the mean body mass in each group:

$$\text{Group A:} \quad \E[Y_A] = \beta_0 + \beta_1*201$$
$$\text{Group B:} \quad \E[Y_B] = \beta_0 + \beta_1*200$$

Then take the difference between them:
\begin{align*}
\E[Y_A] - \E[Y_B] &= \left(\beta_0 + \beta_1*201 \right) - \left(\beta_0 + \beta_1*200\right)\\
& = 201\beta_1 - 200\beta_1\\
&= \beta_1
\end{align*}

Possible summarizing sentences for $\beta_1$:

* $\beta_1$ is the difference in average body mass (in g) for penguins that differ in flipper length by 1 mm. 
* The average difference in body mass for penguins that have 1 mm longer flippers is $\beta_1$ grams.
* The difference in average body mass for penguins that differ in flipper length by one millimeter is $\beta_1$ grams.


```{example}
In Example \@ref(exm:peng-beta1), what is the interpretation of $10\beta_1$?
```

### Interpreting $\beta_0$


The $\beta_0$ parameter represents the **intercept** of the regression line. In other words, $\beta_0$ is the average value of $Y_i$ (i.e. $\E[Y_i]$) for observations with an $x$ value of 0. While mathematically useful, $\beta_0$ might make no practical or scientific sense!

When describing the interpretation of $\beta_0$, follow all of the same guidelines as for $\beta_1$ above, with the following exception:

* Interpret it as the average **value** of the outcome variable (as opposed to a difference in the average value).
* Specify it corresponds to a value of 0 for $x$, rather than a 1-unit difference in $x$.


\textcolor{red}{TO ADD: A 10-unit difference here}


```{example peng-beta0}
Consider a third group of penguins:
```

* Penguins with flipper length of 0 mm. Call this "Group C"

$$\text{Group C:} \quad \E[Y_C] = \beta_0 + \beta_1*0 = \beta_0$$

Equivalent interpretation statements for $\beta_0$ in this context:

* $\beta_0$ is the average body mass (in g) for penguins that have a flipper length of 0 mm. 
* The average body mass for penguins that have 0 mm long flippers is $\beta_0$ grams.

### Interpreting $\beta_0 + \beta_1x_i$

From Equation \@ref(eq:slrmean), the points on the regression line can be interpreted as the average value of the outcome variable for units with a particular value of $x$.

```{example}
Consider penguins in Group A (from Example \@ref(exm:peng-beta1)). In terms of parameters in the simple linear regression mdoel, what is their average body mass?
```  

Since penguins in Group A have flipper lengths of $x_i = 200$mm, their average body mass is given by:

$$\E[Y_i | x_i = 200] = \beta_0 + 200\beta_1.$$

In Section \@ref(slrCIMean), we will see more on how to interpret this value and the difference between estimating a mean and predicting a new observation.


### Interpretation of $\sigma^2$

The parameter $\sigma^2$ represents the variance of the data around the regression line:

\begin{align*}
\Var(Y_i) &=  \Var(\beta_0 + \beta_1x_i + \epsilon_i)\\
&=  \Var(\beta_0 + \beta_1x_i) + \Var(\epsilon_i)\\
&= 0 + \sigma^2\\
&= \sigma^2
\end{align*}

* A large value of $\sigma^2$ means that data are more spread out vertically around the line.
* A small value of $\sigma^2$ means that data are  vertically close to the line


The following two plots show simulated data. The left panel has data generated from a model with $\sigma^2 = 10$ and the right panel data comes from a model with $\sigma^2 = 1$.

```{r echo=FALSE}
n <- 100
beta0 <- 1
beta1 <- 2 
set.seed(11)
x <- runif(n=n, 0, 2)
y1 <- beta0 + beta1*x + rnorm(n, sd=sqrt(10))
y2 <- beta0 + beta1*x + rnorm(n, sd=sqrt(1))
simdata <- data.frame(x=c(x, x),
                      y=c(y1, y2),
                      sigma2=rep(c(10, 1), each=n))
simdata$sigma2lab <- factor(simdata$sigma2,
                            levels=c(10, 1),
                            labels=c(expression(sigma^2*" = 10"), expression(sigma^2*" = 1")))
```

```{r echo=FALSE, fig.cap="Simulated data showing the impact of different values of $\\sigma^2$."}
ggplot(simdata) + theme_bw() + 
  geom_point(aes(x=x, y=y)) +
  geom_abline(aes(intercept=1, slope=2)) +
   facet_wrap(~sigma2lab, labeller=label_parsed)
```


## Random Variables v. Data {#sec:realtheo}

The SLR model \@ref(eq:slr) is an equation for the line for theoretical random variables $Y_i$. In practice, we observed data $y_i$ and estimate a line that has an estimated slope and intercept.


In other words, real data are not generated from the theoretical model 
$$Y_i = \beta_0 + \beta_1x_i + \epsilon_i$$
But for a dataset with specific values $(x_i, y_i)$, we can use the theoretical model to describe the data:

\begin{equation}
y_i = \hat\beta_0 + \hat\beta_1x_i + e_i 
(\#eq:slrest)
\end{equation}

In equation \@ref(eq:slrest):

* $y_i$ is the *observed* value of the outcome for observation $i$
* $e_i$ is the *residual* value corresponding to observation $i$
* $\hat\beta_0$ is the *estimated* intercept for the regression line
* $\hat\beta_1$ is the *estimated* slope for the regression line
* $x_i$ is observed value of the predictor variable for observation $i$

The "hats" in $\hat\beta_0$ and $\hat\beta_1$ indicate that the values are estimates of the true parameter $\beta_1$ and $\beta_0$. The difference between the theoretical model and estimated model can be summarized in the following table:

| Theoretical/Math World |  Real World Data |
|:--------|:---------|
| $Y_i =$ outcome (a random variable) | $y_i =$ outcome (a known number) |
| $x_i =$ predictor (a known number) | $x_i =$ predictor (a known number) |
| $\epsilon_i=$ error (a random variable) | $e_i =$ residual (a known number) |
| $\beta_0 =$ intercept parameter (an unknown number) | $\hat\beta_0 =$ intercept *estimate* (a calculated number) |
| $\beta_1 =$ slope parameter (an unknown number) | $\hat\beta_1 =$ slope *estimate* (a calculated number)  |
| $\sigma^2 =$ variance parameter (an unknown number)  | $\hat\sigma^2 =$ variance *estimate* (a calculated number) |

## Examples 

<!-- Penguin Data -->

```{example}
In the example of penguin flipper length and body mass, the equation for the estimated regression line is:

$$y_i = -5780.8 + 49.7*x_i$$
We can interpret this line in the following sentence:

*A difference of one mm in flipper length is associated with a difference of 49.7 g greater average body mass among penguins in Antarctica.*
```

Note the key elements in this sentence:

* "average" -- Linear regression model tells us about the mean, not a specific observation
* "one mm in flipper length" -- always provide units
* "among penguins in Antarctica" -- Context and/or population
* Observational study -- association not causation!

Figure \@ref(fig:g-penguin-labelled) shows the slope and intercept graphically, while Figure \@ref(fig:g-penguin-resid) shows a graphical representation of the residuals for each observation.



```{r g-penguin-labelled, echo=FALSE, message=FALSE, fig.cap="Fitted regression line for penguin data."}
g_penguin_flip_mass_lm + 
  geom_abline(aes(slope=49.7, intercept=-5780.8)) + 
  geom_text(aes(x=10, y=-5780.8, label="y-intercept = -49.7"), hjust=0, vjust=0.5) +
  geom_text(aes(x=75, y=-5780.8 + 75*50, label="Slope= -5,790.8"), hjust=0, vjust=-0.5, angle=39) +
  coord_cartesian(ylim=c(-6500, max(penguins$body_mass_g)),
                  xlim=c(-5, max(penguins$flipper_length_mm)))
```

```{r g-penguin-resid, echo=FALSE, message=FALSE, fig.cap="Residuals for penguin data."}
penguin_lm <- lm(body_mass_g ~ flipper_length_mm, data=penguins)
pen2 <- data.frame(x=penguin_lm$model$flipper_length_mm,
                   obs=penguin_lm$model$body_mass_g,
                   fit=fitted(penguin_lm),
                   id=1:nobs(penguin_lm)) %>%
  pivot_longer(-c(x, id))
g_penguin_flip_mass_lm + 
  geom_line(aes(x=x, y=value, group=id), data=pen2, col="purple")
```




```{r include=FALSE}
lm_penguin <- lm(body_mass_g~flipper_length_mm, data=penguins)
coef(lm_penguin)
```


<!-- ### $\hat\beta_1$: Estimator or Estimate? -->

<!-- $\hat\beta_1$ can represent two things: -->

<!-- * An estimate of $\beta_1$ -->
<!--    * A number -->
<!--    * Interpreted as the **estimated** average difference in $y$ for a one-unit difference in $x$ -->
<!-- * An estimator for $\beta_1$ -->
<!--    * A random variable -->
<!--    * This is a rule for calculating a number that depends on $x$ and $y$ -->

<!-- \vspace{0.5cm} -->

<!-- Why does the difference matter? -->

<!-- * $\hat\beta_1$ depends on the data -->
<!-- * $\hat\beta_1$ has uncertainty (variance and possibly bias) -->

<!-- Note: $\hat\beta_0$ and $\hat\beta_1$ are calculated as numbers, but we need to think of them as realizations of a random generating process. This is because they are functions of the data, which we think of as a realization of a random generating process. -->


## Exercises

```{exercise}
In the context of Example \@ref(exm:peng-beta1), what is an interpretation of $50\beta_1$?
```

```{exercise}
In the context of Example \@ref(exm:peng-beta0), does the quantity $10\beta_0$ make practical sense? If yes, provide an interpretation. If no, explain why not.
```

```{exercise}
Consider a variation of Example \@ref(exm:peng-beta1), in which a simple linear regression mdoel is fit with body mass (in g) as the *predictor* variable and flipper length (in mm) as the outcome. Explain how this switch affects the interpretations of $\beta_0$, $\beta_1$, and $\sigma^2$.
```
